Computational Prediction of Boundaries of 3D Genomic Domains in Class Imbalance Settings

### 2019-08-25 Manuscript/manuscript.Rmd
edits

We analyzed publically available replicate Hi-C data sets from the GM12878 and K562 cell lines (Supplementary Table 1). Topologically associating domains (TADs) were subsequently called using the ARROWHEAD pipeline [@durand2016juicer] at 5 data resolutions: 5 kb 10 kb, 25 kb, 50 kb, and 100 kb. After removing TADs that exceeded 2 megabases in length, we obtained a total of 16133, 14824, 8081, 3560, and 919 unique TAD boundaries for the GM12878 cell line, and 9314, 10881, 6923, 3197, and 831 unique TAD boundaries for the K562 cell line across each of the 5 resolutions respectively (Supplementary Figure 1). To account for the variability in boundary location due to varying resolution, we developed a resolution-specific approach toward constructing TAD versus non-TAD boundary regions on the linear genome (See Methods), labeled as a response vector **Y** (Figure 1). TAD boundary regions were defined as regions on the linear genome where a genomic bin overlapped with a flanked TAD boundary ($Y = 1$), while non-TAD boundary regions were defined as the regions that did not overlap with a flanked TAD boundary ($Y = 0$). A significant number of TAD boundary regions were found to be specific to one of the two cell lines (Supplementary Figure 2). Given the plasticity in boundary location, we opted to build cell line specific models for predicting TAD boundary regions. The results presented are for the GM12878 cell line, with K562 results presented as supplementary.

![**An illustration of the proposed approach for resolution-specific data construction used for TAD boundary prediction.** Unique boundaries are denoted by the blue dots. The yellow regions denote the flanked boundary regions according to the Hi-C data resolution (i.e. 100 kb resolution corresponded to a 100 kb flank for a total boundary region of 200 kb). The linear genome was subsequently binned into resolution-specific intervals. The response vector **Y** was defined as a variable that indicated whether a genomic bin overlapped with a flanked TAD boundary ($Y = 1$) or not ($Y = 0$).](Figures/figure1_model_construction_V7.png)

@@

We then investigated the proportional class imbalance (PI) featured at each resolution and among each cell line. The PI was defined as twice the ratio of TAD boundary regions ($Y=1$) versus the sum of both TAD and non-TAD boundary regions. The proportional imbalance was found to be invariant to resolution, remaining consistently low for both cell lines, with very little variability between chromosomes (Figure 3C). Such low proportions of imbalance gave precedence for the implementation of re-sampling prior to classification.

![**Cell-line specific boundary summaries.** (A) Barplots illustrating that the total number of TAD boundary regions decreases as Hi-C data resolution decreases. (B) Venn diagrams showing that TAD boundary regions exhibit cell-line specificity across resolution. (C) Boxplots indicating that the proportion of class imbalance remained severe irrespective of resolution.](Figures/bounds_and_pis_v2.png)

@@

A total of 78 cell type specific functional genomic annotations derived from ChIP-Seq experiments (See Methods) for the GM12878 and K562 cell line were used to build the feature space of downstream predictive models (Supplementary Tables 2-3). We constructed three different predictor types for each functional genomic annotation. Counts of overlaps (**overlap count**, OC) were defined as the total number of peak regions that overlapped within a given genomic bin (Figure 3A). Percents of overlaps (**overlap percent**, OP) were defined as the proportion of a bin overlapped by a peak region (Figure 3B). **Distance** was defined as log2 transformed number of base pairs between between the centers of bins and regions defining genomic annotations (Figure 3C). A log2-transformation of distance was used to account for the skewness present in the distributions of distances <!--(Supplementary Figures 1-2)-->. Each predictor type provides a granular way to draw associations between functional genomic annotations and the spatial structure of the genome.

![**Three types of associations between TAD boundaries and genomic annotations.** (A) _Overlap Counts_ (OC) consider the total number of genomic features that overlapped with each genomic bin. (B) _Overlap Percent_ (OP) was calculated by dividing the sum of all feature widths within each specific bin by the total bin width. (C) _Distance_ ($log_{2}$ transformed) was calculated by measuring the distance (in bases) from the center of each genomic bin to the center of the nearest region defining a genomic annotation. ](Figures/figure2_predictor_types.png)

>>>>

We analyzed publically available replicate Hi-C data sets from the GM12878 and K562 cell lines (Supplementary Table 1). Topologically associating domains (TADs) were subsequently called using the ARROWHEAD pipeline [@durand2016juicer] at 5 data resolutions: 5 kb 10 kb, 25 kb, 50 kb, and 100 kb. After removing TADs that exceeded 2 megabases in length, we obtained a total of 16133, 14824, 8081, 3560, and 919 unique TAD boundaries for the GM12878 cell line, and 9314, 10881, 6923, 3197, and 831 unique TAD boundaries for the K562 cell line across each of the 5 resolutions respectively (Supplementary Figure 1). To account for the variability in boundary location due to varying resolution, we developed a resolution-specific approach toward constructing TAD versus non-TAD boundary regions on the linear genome (See Methods), labeled as a response vector **Y** (Figure 1A). TAD boundary regions were defined as regions on the linear genome where a genomic bin overlapped with a flanked TAD boundary ($Y = 1$), while non-TAD boundary regions were defined as the regions that did not overlap with a flanked TAD boundary ($Y = 0$). A significant number of TAD boundary regions were found to be specific to one of the two cell lines (Supplementary Figure 2). Given the plasticity in boundary location, we opted to build cell line specific models for predicting TAD boundary regions. The results presented are for the GM12878 cell line, with K562 results presented as supplementary.

![**Illustrating the resolution-specific data construction and feature engineering used for downstream modelling.** (A) Unique boundaries are denoted by the blue dots. The yellow regions denote the flanked boundary regions according to the Hi-C data resolution (i.e. 100 kb resolution corresponded to a 100 kb flank for a total boundary region of 200 kb). The linear genome was subsequently binned into resolution-specific intervals. The response vector **Y** was defined as a variable that indicated whether a genomic bin overlapped with a flanked TAD boundary ($Y = 1$) or not ($Y = 0$). (B) Three types of associations between TAD boundaries and genomic annotations including _Overlap Counts_ (OC), _Overlap Percent_ (OP), and ($log_{2}$) _Distance_. The green shapes denote regions defined by a specific functional genomic annotation used for feature engineering. ](Figures/model_construction_and_predictor_types.png)

@@

We then investigated the proportional class imbalance (PI) featured at each resolution. The PI was defined as the ratio of TAD boundary regions ($Y=1$) versus the sum of both TAD and non-TAD boundary regions for each chromosome specific sections of the linear genome and has a range from 0 to 0.5. The PI was found to be invariant to resolution, remaining consistently low, with an average of 0.06, 0.10, 0.14, 0.12, and 0.06 for the 5 kb, 10 kb, 25 kb, 50 kb, and 100 kb resolutions respectively. The K562 cell line was found to have comparatively low proportional imbalances (Supplementary Figure 3). Such low proportions of imbalance gave precedence for the implementation of re-sampling prior to performing classification.

@@

A total of 78 cell type specific functional genomic annotations derived from ChIP-Seq experiments (See Methods) for the GM12878 and K562 cell line were used to build the feature space of downstream predictive models (Supplementary Tables 2-3). We constructed three different predictor types for each functional genomic annotation. Counts of overlaps (**overlap count**, OC) were defined as the total number of peak regions that overlapped within a given genomic bin (Figure 1B). Percents of overlaps (**overlap percent**, OP) were defined as the proportion of a bin overlapped by a peak region (Figure 3B). **Distance** was defined as log2 transformed number of base pairs between between the centers of bins and regions defining genomic annotations (Figure 3C). A log2-transformation of distance was used to account for the skewness present in the distributions of distances <!--(Supplementary Figures 1-2)-->. Each predictor type provides a granular way to draw associations between functional genomic annotations and the spatial structure of the genome.
